\newif\ifieeeaccessclass
\IfFileExists{ieeeaccess.cls}{
\ieeeaccessclasstrue
\documentclass{ieeeaccess}
}{
\ieeeaccessclassfalse
\documentclass[10pt]{article}
}

\ifieeeaccessclass\else
\newcommand{\history}[1]{}
\newcommand{\doi}[1]{}
\newcommand{\address}[1]{}
\newcommand{\corresp}[1]{}
\newenvironment{keywords}{\par\noindent\textbf{Keywords: }\itshape}{\par}
\fi

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\emergencystretch=4pt
\usepackage{tikz}
\usetikzlibrary{positioning,shapes.geometric}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{longtable}

        \usepackage{longtable}
        \usepackage{pdflscape}
        \usepackage{geometry}

  \usepackage{colortbl}   % for \rowcolor and \cellcolor
  \usepackage{xcolor}     % for \definecolor and color names
  \usepackage{array}      % for extended column formatting
  \usepackage{booktabs}   % already present in v6
  \usepackage{multirow}   % for \multirow in phase label column


  \usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{array}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}


\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{tikz}
\usetikzlibrary{
    positioning,
    fit,
    arrows.meta,
    shapes.geometric,
    shapes.misc,
    calc
}



\history{Received XX Month XXXX, revised XX Month XXXX, accepted XX Month XXXX.}
\doi{10.1109/ACCESS.XXXX.DOI}

\title{Cybersecurity \& Cyber-Resilience Strategy for the Sovereign Cloud--Edge--IoT Continuum}
\author{Iraklis Symeonidis}
\address{RISE Research Institutes of Sweden, Sweden}
\corresp{Corresponding author: Iraklis Symeonidis (e-mail: iraklis.symeonidis@ri.se).}

\begin{document}

\hypersetup{pageanchor=false}
\maketitle
\hypersetup{pageanchor=true}

\begin{abstract}
The rapid decentralization of computing infrastructures toward the Cloud--Edge--IoT continuum introduces unprecedented cybersecurity challenges, including expanded attack surfaces, heterogeneous environments, and increased dependency on non-sovereign technologies. This paper proposes a structured, multi-stage methodology for defining and prioritizing cybersecurity capabilities required to achieve a secure and sovereign European computing continuum. The approach combines a pipeline-oriented transformation model with a seven-dimensional analytical framework, enabling traceable transitions from architectural capabilities to strategic priorities and policy-aligned recommendations. A four-filter prioritization logic identifies critical capabilities based on regulatory urgency, architectural dependencies, sovereignty risks, and market feasibility. These priorities are further analyzed through a progression matrix capturing operational deployment and strategic sovereignty evolution across time phases. The methodology provides a reproducible and evidence-based framework for guiding research, innovation investments, and policy decisions in the context of the European digital sovereignty agenda.
\end{abstract}

\begin{keywords}
Cybersecurity, Cloud-Edge-IoT, Digital Sovereignty, Zero Trust, Post-Quantum Cryptography, EU Policy, Cyber-Resilience
\end{keywords}

\include{1.introduction}
 
\include{2.Related_work}

\newpage
\include{Methodology}


\newpage
\include{capabilities}

\newpage
\include{consolidated_priorities_v1}


% =========================================================================
% SECTION: Executive Summary & Strategic Recommendations
% Compressed version — three-stream structure, priority-anchored,
% gate-traceable, phase-differentiated verb register.
% Applied R&D exceptions stated once per phase preamble.
% Success metrics retained verbatim.
% =========================================================================

% \include{9.Executirve_summary}
\newpage

\newpage
% =========================================================================
% CHAPTER 2: PROMPT
% =========================================================================
\appendix

\newpage
% \include{Discussions}  

\newpage
% \include{Conclusions}

% =========================================================================
% REFERENCES
% =========================================================================
% Keep bibliography generation valid even when citation-heavy sections are commented.
\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{references.bib}

\end{document}